---
title: "Bioinformagician tutorial"
author: "Hannah Bazin"
output: html_document
date: "2025-01-10"
---

This is a script to manipulate gene expression data. The directory is the following:
# setwd("~/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/bioinformagician_learning")

# Set up

## Load libraries

```{r}
library(dplyr)
library(tidyverse)
library(GEOquery)
```

## Load data

```{r}
dat <- read.csv(file = "GSE183947_fpkm.csv")
dim(dat)
```

## Load metadata

```{r}
gse <- getGEO(GEO = "GSE183947", GSEMatrix = TRUE)
```

```{r}
metadata <- pData(phenoData(gse[[1]]))
head(metadata)
```
# Processing

## Process data

Filter the columns in the metadata.

```{r}
metadata_modified <- metadata %>% 
  # keep only relevant columns
  select(c(1, 10, 11, 17)) %>% 
  # rename columns (new_name = old_name)
  rename(tissue = characteristics_ch1) %>% 
  rename(metastasis = characteristics_ch1.1) %>% 
  # replace "tissue: " with nothing
  mutate(tissue = gsub("tissue: ", "", tissue)) %>% 
  mutate(metastasis = gsub("metastasis: ", "", metastasis))
```

Now we can link this metadata to the gene expression data.

```{r}
head(dat)
```

This data is in the wide format (genes as columns), but it is easier to add the metadata if the data is in long format (genes as rows).

Reshape the data.

```{r}
dat_long <- dat %>% 
  rename(gene = X) %>% 
  # convert wide format to long format
  # key = name of the column that will contain the old column names
  # do not change the gene column
  gather(key = "samples", value = "FPKM", -gene)
```

Join dataframes = dat_long + metadata_modified.

```{r}
dat_long <- dat_long %>% 
  left_join(., metadata_modified, by = c("samples" = "description"))
```

## Explore data

Extract the expression for BRCA1 and BRCA2, and compare the expression between normal and tumour samples.

```{r}
dat_long %>% 
  filter(gene == "BRCA1" | gene == "BRCA2") %>% 
  group_by(gene, tissue) %>% 
  # create a new column with the mean FPKM
  summarize(mean_FPKM = mean(FPKM),
            median_FPKM = median(FPKM)) %>% 
  # sort the column in the ascending order (add minus for descending order: -mean_FPKM)
  arrange(-mean_FPKM)
```

## Visualise data

Basic format for ggplot

```{r}
ggplot(data, aes(x = variable, y = variable1)) +
  geom_col()
```

1. Barplot

```{r}
dat_long %>% 
  filter(gene == "BRCA1") %>% 
  ggplot(., aes(x = samples, y = FPKM, fill = tissue)) +
  geom_col()
```

2. Density

See if BRCA1 expression is different in breast tumour tissue or normal breast tissue.

```{r}
dat_long %>% 
  filter(gene == "BRCA1") %>% 
  ggplot(., aes(x = FPKM, fill = tissue)) +
  geom_density(alpha = 0.3)
```

3. Boxplot

See if there is a difference in BRCA1 expression in metastasis vs no metastasis.

```{r}
dat_long %>% 
  filter(gene == "BRCA1") %>% 
  ggplot(., aes(x = metastasis, y = FPKM)) +
  #geom_boxplot()
  geom_violin()
```

4. Scatterplot

Compare the expression of two genes.


```{r}
dat_long %>% 
  filter(gene == "BRCA1" | gene == "BRCA2") %>%
  # convert from long to wide format
  spread(key = gene, value = FPKM) %>% 
  ggplot(., aes(x = BRCA1, y = BRCA2, colour = tissue)) +
  geom_point() +
  # add a straight line
  geom_smooth(method = "lm", se = FALSE)
```

5. Heatmap

Visualise multiple genes across samples and compare the expression in all samples.

```{r}
genes_of_interest <- c("BRCA1", "BRCA2", "TP53", "ALK", "MYCN")

dat_long %>% 
  filter(gene %in% genes_of_interest) %>% 
  ggplot(., aes(x = samples, y = gene, fill = FPKM)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red")
```

Save the plot.

First way:

```{r}
p <- dat_long %>% 
  filter(gene %in% genes_of_interest) %>% 
  ggplot(., aes(x = samples, y = gene, fill = FPKM)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red")

ggsave(p, filename = "heatmap_save1.pdf", width = 10, height = 8)
```

Second way (does not work as of now):

```{r}
pdf("heatmap_save2.pdf", width = 10, height = 8)

dat_long %>% 
  filter(gene %in% genes_of_interest) %>% 
  ggplot(., aes(x = samples, y = gene, fill = FPKM)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red")

dev.off()
```










