
```{r}

library(GEOquery)
library(Seurat)
library(dplyr)
library(patchwork)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(pathfindR)

```

## Bar plots

Load data and list patient conditions.

```{r}

# Load data
early_df <- read_csv("results/humanPVATsn/pathfindR/comparison1v3/early_pre_healthy_vs_obese.csv")
intermediate_df <- read_csv("results/humanPVATsn/pathfindR/comparison1v3/intermediate_pre_healthy_vs_obese.csv")
mature_df <- read_csv("results/humanPVATsn/pathfindR/comparison1v3/mature_healthy_vs_obese.csv")

# List of conditions
conditions <- c("A only", "B only", "common")
condition_labels <- list("A only" = "Healthy Only", "B only" = "Obese Only", "common" = "Both Healthy and Obese")

```

Define function to create and save bar plots.
- For the A only and B only pathways, we rank them based on fold enrichment.
- For the common pathways, we rank them based on the combined p-value; this helps identify pathways that are consistently significant in both datasets.

```{r}

create_and_save_bar_plot <- function(df, category, title_label, output_file) {
  
  # Define sorting column
  sorting_col <- ifelse(category == "common", "combined_p", paste0("Fold_Enrichment_", substr(category, 1, 1)))

  # Filter top 20 enriched pathways
  if(category == "common") {
    df_filtered <- df %>% 
      filter(status == category) %>%
      arrange(combined_p) %>%
      slice_head(n = 20)
  } else {
    df_filtered <- df %>% 
      filter(status == category) %>%
      arrange(desc(.data[[sorting_col]])) %>%
      slice_head(n = 20)
  }

  # Ensure factor levels are correctly ordered for plotting
  if(category == "common") {
     df_filtered <- df_filtered %>%
       mutate(Term_Description = factor(Term_Description,
                                        levels = Term_Description))
  } else {
    df_filtered <- df_filtered %>%
      mutate(Term_Description = factor(Term_Description, levels = rev(Term_Description)))

  }
  
  # Create bar plot
  plot <- ggplot(df_filtered, aes(x = .data[[sorting_col]], y = Term_Description, fill = Term_Description)) +
    geom_bar(stat = "identity", color = "black") +
    labs(title = title_label, x = ifelse(category == "common", "Combined P Value", "Fold Enrichment"),
         y = "Pathway") +
    theme_minimal() +
    theme(legend.position = "none")
  
  
  # Save plot
  ggsave(output_file, plot, width = 12, height = 6, device = "pdf")

  # Return plot
  return(plot)
  
}

```

Generate plots for each differentiation stage: early pre-adipocytes, intermediate pre-adipocytes, and mature adipocytes.

```{r fig.width=12, fig.height=6}

for (condition in conditions) {
  
  # Early adipocytes
  early_plot <- create_and_save_bar_plot(early_df, condition, paste("Early Pre-adipocytes: Enriched Pathways for", condition_labels[[condition]]), paste0("results/humanPVATsn/pathfindR/comparison1v3/early_", gsub(" ", "_", condition), ".pdf"))
  print(early_plot)

  # Intermediate adipocytes
  inter_plot <- create_and_save_bar_plot(intermediate_df, condition, paste("Intermediate Pre-adipocytes: Enriched Pathways for", condition_labels[[condition]]), paste0("results/humanPVATsn/pathfindR/comparison1v3/intermediate_", gsub(" ", "_", condition), ".pdf"))
  print(inter_plot)
  
  # Mature adipocytes
  mature_plot <- create_and_save_bar_plot(mature_df, condition, paste("Mature Adipocytes: Enriched Pathways for", condition_labels[[condition]]), paste0("results/humanPVATsn/pathfindR/comparison1v3/mature_", gsub(" ", "_", condition), ".pdf"))
  print(mature_plot)
  
}

```