---
title: "Learning OmnipathR"
author: "Alberto Valdeolivas, Attila Gabor, Denes Turei, Julio Saez-Rodriguez"
output: html_document
date: "2024-12-04"
---

# Setup

Load libraries.

```{r}
library(OmnipathR)
library(igraph)
library(tidyr)
library(gprofiler2)
```

# Usage examples

## Interactions

Check source databases.

```{r}
## We check some of the different interaction databases
get_interaction_resources()
```
View interactions (from OmniPath dataset).

```{r}
## The interactions are stored into a data frame.
pathways <- import_omnipath_interactions(resources = c("SignaLink3", "PhosphoSite", "SIGNOR"))

## We visualize the first interactions in the data frame.
print_interactions(head(pathways))
```

Convert these interactions into graph networks.

```{r}
## We transform the interactions data frame into a graph
pathways_g <- interaction_graph(pathways)
```

### Shortest path

Finding a path between two nodes (proteins) going through the minimum number of edges.

Two similar functions: print_path_es takes as input an edge sequence; print_path_vs takes as input a node sequence.

```{r}
## Find and print shortest paths on the directed network between proteins of interest:
print_path_es(
  igraph::shortest_paths(
    pathways_g,
    from = "TYRO3",
    to = "STAT3",
    output = "epath"
    )$epath[[1]],
  pathways_g
)
```

```{r}
## Find and print all shortest paths between proteins of interest:
print_path_vs(
  igraph::all_shortest_paths(
    pathways_g,
    from = "DYRK2",
    to = "MAPKAPK2"
  )$res,
  pathways_g
)
```

### Clustering

Grouping nodes (proteins) so that nodes belonging to the same cluster are more connected in the network to each other than those in other clusters. Proteins in the same cluster are more likely to be implicated in similar biological tasks.

We apply a clustering algorithm (Louvain) to group proteins in our network. We apply here Louvain which is fast but can only run on undirected graphs. Other clustering algorithms can deal with directed networks but with longer computational times, such as cluster_edge_betweenness. These cluster methods are directly available in the igraph package.

```{r}
## Apply a clustering algorithm
pathways_g_u <- igraph::as_undirected(pathways_g, mode = "mutual")
pathways_g_u <- igraph::simplify(pathways_g_u)
clusters <- igraph::cluster_fast_greedy(pathways_g_u)
```

Extract the cluster where a protein of interest is contained.

```{r}
erbb2_cluster_id <- clusters$membership[which(clusters$names == "ERBB2")]
erbb2_cluster_g <- igraph::induced_subgraph(
  pathways_g_u,
  igraph::V(pathways_g)$name[which(clusters$membership == erbb2_cluster_id)]
)
```


## Other interactions

We can import the datasets from different sources than OmniPath. The same functions can be applied to every interaction dataset.

### Pathway Extra

This dataset contains activity flow interactions without literature reference. We are focusing on mouse interactions for a given gene.

```{r}
# Query and store interactions into a dataframe
iactions <- import_pathwayextra_interactions(
  resources = c("Wang", "Lit-BM-17"),
  organism = 10090 # mouse
)
```

```{r}
# Select all interactions in which Amfr gene is involved
iactions_amfr <- dplyr::filter(
  iactions,
  source_genesymbol == "Amfr" |
  target_genesymbol == "Amfr"
)
# Print these interactions
print_interactions(iactions_amfr)
```

### Kinase Extra

This dataset contains enzyme-substrate interactions without literature reference. We focus on rat reactions targeting a particular gene.

```{r}
# Query and store interactions into a dataframe
phosphonetw <-
  import_kinaseextra_interactions(
    resources = c("PhosphoPoint", "PhosphoSite"),
    organism = 10116 # rat
  )
```

```{r}
# Select interactions in which Dpysl2 gene is a target
upstream_dpysl2 <- dplyr::filter(
  phosphonetw,
  target_genesymbol == "Dpysl2"
)
# Print these interactions
print_interactions(upstream_dpysl2)
```

### Ligand-receptor Extra

This dataset contains ligand-receptor interactions without literature reference. We want to find the potential receptors associated to a given ligand, CDH1.

```{r}
# Query and store interactions into a dataframe
ligrec_netw <- import_ligrecextra_interactions(
  resources = c("iTALK", "Baccin2019"),
  organism = 9606 # human
)
```

```{r}
# Receptors of CDH1 ligand
downstream_cdh1 <- dplyr::filter(
  ligrec_netw,
  source_genesymbol == "CDH1"
)

# Transform dataframe into a graph
downstream_cdh1_g <- interaction_graph(downstream_cdh1)

# Induce a network with these genes
downstream_cdh1_g <- igraph::induced_subgraph(
  interaction_graph(import_omnipath_interactions()),
  igraph::V(downstream_cdh1_g)$name
)
```


### DoRothEA Regulons

This dataset contains TF-target interactions with confidence score, ranging from A-E (A is the most confident).

We can select and print the most confident interactions for a given TF.

```{r}
# Query and store the interactions into a dataframe
dorothea_netw <- dorothea(
  dorothea_levels = "A",
  organism = 9606 # human
)
```

```{r}
downstream_gli1 <- dplyr::filter(
  dorothea_netw,
  source_genesymbol == "GLI1"
)
print_interactions(downstream_gli1)
```

### miRNA-target database

This dataset stores miRNA-mRNA and TF-miRNA interactions. These are only available for human so far. We select the miRNA interacting with the TF GLI1. The main function of miRNAs seems to be related with gene regulation. It is therefore interesting to see how some miRNA can regulate the expression of a TF which in turn regulates the expression of other genes.

```{r}
## We query and store the interactions into a dataframe
mirna_target <- import_mirnatarget_interactions(
    resources = c("miR2Disease", "miRDeathDB")
)
```


```{r}
## We select the interactions where a miRNA is interacting with the TF
## used in the previous code chunk and we print these interactions.
upstream_gli1 <- dplyr::filter(
    mirna_target,
    target_genesymbol == "GLI1"
)

print_interactions(upstream_gli1)
```

Transform into graphs.

```{r}
## We transform the previous selections to graphs (igraph objects)
downstream_gli1_g <- interaction_graph(downstream_gli1)
upstream_gli1_g <- interaction_graph(upstream_gli1)
```


### Small molecule-protein dataset

This is a small recently added dataset (2022, 3.5k interactions) from three resources (SIGNOR, CancerDrugsDB and Cellinker).

We can look for targets of a cancer drug, the MEK inhibitor Trametinib.

```{r}
trametinib_targets <- import_small_molecule_protein_interactions(
    sources = "TRAMETINIB"
)
print_interactions(trametinib_targets)
```

Note, the human readable compound names are not reliable, use PubChem CIDs instead.

## Post-translational modifications (PTMs)

In the next code chunk, we download the PTMs for human. We first check the different available source databases, even though we do not perform any filter. Then, we select and print the reactions involving a specific enzyme-substrate pair. Those reactions lack information about activation or inhibition. To obtain that information, we match the data with OmniPath interactions. Finally, we show that it is also possible to build a graph using this information, and to retrieve PTMs from mouse or rat.

```{r}
# Check different PTM databases
get_enzsub_resources()
```
```{r}
# Query and store the enzyme-PTM interactions into a dataframe, no filtering in this case
enzsub <- import_omnipath_enzsub()

# Select and print interactions between specific kinase and substrate
print_interactions(dplyr::filter(
  enzsub,
  enzyme_genesymbol == "MAP2K1",
  substrate_genesymbol == "MAPK3"
))
```
In the previous results, we can see that enzyme-PTM relationships do not contain sign (activation/inhibition). We can generate this information based on the protein-protein OmniPath interaction dataset.

```{r}
interactions <- import_omnipath_interactions()
enzsub <- get_signed_ptms(enzsub, interactions)

# Select the same kinase and substrate, now print with information about inhibition or activation
print_interactions(dplyr::filter(enzsub,
  enzyme_genesymbol == "MAP2K1",
  substrate_genesymbol == "MAPK3"
))
```

Transform enzyme-PTM relationships into a graph.

```{r}
enzsub_g <- enzsub_graph(enzsub = enzsub)
```

Download PTMs for mouse.

```{r}
enzsub <- import_omnipath_interactions(
  resources = c("PhosphoSite", "SIGNOR"),
  organism = 10090
)
```

## Complexes

The complexes query provides access to more than 20000 protein complexes. This comprehensive database has been created by integrating different resources. We now download these molecular complexes filtering by some of the source databases. We check the complexes where a couple of specific genes participate. First, we look for the complexes where any of these two genes participate. We then identify the complex where these two genes are jointly involved. Finally, we perform an enrichment analysis with the genes taking part in that complex. You should keep an eye on this complex since it will be used again in the forthcoming sections.


```{r}
# Check the different complex databases
get_complex_resources()
```
```{r}
# Query and store complexes from some sources into a dataframe¨
complexes <- import_omnipath_complexes(resources = c("CORUM", "hu.MAP"))

# Check all complexes where a set of genes participate
query_genes <- c("WRN", "PARP1")
```


We can select the complexes where any of the input genes participate.

```{r}
# Complexes where any of the input genes participate
wrn_parp1_complexes <- unique(
  get_complex_genes(
    complexes,
    query_genes,
    total_match = FALSE
  )
)

# Print components of the selected complexes
head(wrn_parp1_complexes$components_genesymbols, 6)
```


Or we can choose complexes where all the input genes participate jointly.

```{r}
complexes_query_genes_join <- unique(
  get_complex_genes(
    complexes,
    query_genes,
    total_match = TRUE
  )
)

# Print components of the selected complexes
complexes_query_genes_join$components_genesymbols
```

We can perform an enrichment analysis with the genes in the complex.

GO: gene ontology
BP: biological process
CC: cellular component
MF: molecular function

```{r}
wrn_parp1_cplx_genes <- unlist(
  strsplit(wrn_parp1_complexes$components_genesymbols, "_")
)

# Enrichment analysis with the genes in the complex
enrichment <- gprofiler2::gost(
  wrn_parp1_cplx_genes,
  significant = TRUE,
  user_threshold = 0.001,
  correction_method = "fdr",
  sources = c("GO:BP", "GO:CC", "GO:MF")
)

# Show the most significant results
enrichment$result %>% 
  dplyr::select(term_id, source, term_name, p_value) %>% 
  dplyr::top_n(5, -p_value)
```

## Annotations

Annotations are statements about the different features of a biological entity. The data come from different databases. There is a record_id field which is unique within the records of each database. Each row has a key-value pair, and you need the record_id to connect the related key-value pairs.

```{r}
# Check different annotation databases
get_annotation_resources()
```

We can further investigate the features of the complex selected in the previous section.

```{r}
# Get the annotations of the complex itself
annotations <- import_omnipath_annotations(
  proteins = paste0("COMPLEX:", wrn_parp1_complexes$components_genesymbols)
)

head(dplyr::select(annotations, source, label, value), 10)
```

We then explore the annotations of the individual components of the complex.

Pathways of every protein in the complex:

```{r}
annotations <- import_omnipath_annotations(
  proteins = wrn_parp1_cplx_genes,
  resources = "NetPath"
)

dplyr::select(annotations, genesymbol, value)
```

Subcellular localisation of our proteins:

```{r}
annotations <- import_omnipath_annotations(
  proteins = wrn_parp1_cplx_genes,
  resources = "ComPPI"
)

# Rearrange the data
annotations %>%
  tidyr::spread(label, value) %>% 
  dplyr::arrange(desc(score)) %>% 
  dplyr::top_n(10, score)
```

We can do the same much easier by passing wide = TRUE as a parameter.

```{r}
signaling_pathways <- import_omnipath_annotations(
  resources = "SignaLink_pathway",
  wide = TRUE
)
```


## Intercell

We can explore intercellular signalling (intercell). This dataset provides information about the roles of proteins in inter-cellular signaling (if it is a ligand, receptor, ECM component, etc.) The data comes from different databases combined into categories.

```{r}
# Check the different intercell categories
head(get_intercell_categories(), 30)
```

```{r}
# Import intercell data into a dataframe
intercell <- import_omnipath_intercell(
  scope = "generic",
  aspect = "locational"
)

# Check the intercell annotations for the individual components of our previous complex
dplyr::filter(intercell, genesymbol %in% wrn_parp1_cplx_genes) %>% 
  dplyr::distinct(genesymbol, parent, .keep_all = TRUE) %>% 
  dplyr::select(category, genesymbol, parent) %>% 
  dplyr::arrange(genesymbol)
```
The import_intercell_network function creates the most complete network, but with many interactions that are false positives. Highly recommended to apply some filtering (with the high_confidence parameter which performs a quiet stringent filtering).

```{r}
icn <- import_intercell_network(high_confidence = TRUE)
```

Or we can use the function filter_intercell_network instead, to have more flexibility to adjust the stringency of the filtering. See function documentation for more detail.

```{r}
icn <- import_intercell_network() %>% 
  filter_intercell_network(
    min_curation_effort = 1,
    consensus_percentile = 33
  )
```

We close graphical connections.

```{r}
while (!is.null(dev.list())) dev.off()
```














