---
title: "Pathway enrichment analysis of targets of significant drugs"
author: "Hannah M. L. Bazin"
date: "r doc_date()"
output: BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
---

# Set up

## Load libraries

```{r}

library(dplyr)
library(gprofiler2)
library(igraph)

```

## Set latest g:Profiler

```{r}

set_base_url("https://biit.cs.ut.ee/gprofiler")

```


## Set working directory

The working directory can be set by running the following command in the console: `setwd("/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project")`.

# Step 1 of differentiation - key genes only - 1000 iterations

## Load data

```{r}

# Drug–target assignments
step1_targets_key_1000_path <- "/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/results/humanPVATsn/network_analysis/proximity_significant_drugs/significant_targets_key_1000_step1.csv"
step1_targets_df <- read.csv(step1_targets_key_1000_path)

# Protein–protein interaction network
combined_PPI_path <- "/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/data/networks/combined_PPI.csv"
ppi <- read.csv(combined_PPI_path)
colnames(ppi) <- c("Gene_A", "Gene_B")

```

## Build PPI graph

```{r}

ppi_graph <- graph_from_data_frame(ppi, directed = FALSE)

get_neighbours <- function(gene) {
  if (gene %in% V(ppi_graph)$name) {
    neighbors(ppi_graph, gene) %>% names()
  } else {
    return(character(0))
  }
}

```

## Collapse targets per drug

```{r}

step1_drug_target_list <- step1_targets_df %>%
  group_by(Drug_Name) %>%
  summarise(Targets = paste(unique(Drug_Target), collapse = ","))

```

## Expand targets for drugs with <6 targets

```{r}

step1_drug_target_list$Expanded_Targets <- NA

for (i in 1:nrow(step1_drug_target_list)) {
  targets <- unlist(strsplit(step1_drug_target_list$Targets[i], ","))
  if (length(targets) < 6) {
    extended <- unique(unlist(c(targets, unlist(lapply(targets, get_neighbours)))))
    step1_drug_target_list$Expanded_Targets[i] <- paste(extended, collapse = ",")
    cat(step1_drug_target_list$Drug_Name[i], "- extended from", length(targets), "to", length(extended), "targets\n")
  } else {
    step1_drug_target_list$Expanded_Targets[i] <- step1_drug_target_list$Targets[i]
  }
}

```

## Run g:Profiler Reactome enrichment per drug

```{r}

# Create folder to save results if it doesn't exist
results_dir <- "/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/results/humanPVATsn/network_analysis/drug_enrichment/step1_key_1000/"
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

for (i in 1:nrow(step1_drug_target_list)) {
  
  drug <- step1_drug_target_list$Drug_Name[i]
  gene_list <- unlist(strsplit(step1_drug_target_list$Expanded_Targets[i], ","))
  
  cat("Running enrichment for:", drug, "-", length(gene_list), "genes\n")
  
  # Run pathway enrichment
  enrichment <- gost(query = gene_list, organism = "hsapiens", sources = "GO:BP", significant = TRUE)
  
  # Check that enrichment worked, then save
  if (!is.null(enrichment) && nrow(enrichment$result) > 0) {
    
    out_file <- paste0(results_dir, gsub("[^a-zA-Z0-9]", "_", drug), ".csv")
    
    # Convert list-columns (like `parents`) to character
    enrichment_clean <- enrichment$result
    enrichment_clean$parents <- sapply(enrichment_clean$parents, function(x) paste(x, collapse = "|"))
    
    # Drop unnecessary column
    enrichment_clean <- enrichment_clean[, !(names(enrichment_clean) %in% c("evidence_codes"))]    
    
    # Save as CSV file
    write.csv(enrichment_clean, out_file, row.names = FALSE)
    
  } else {
    
    cat("No enriched Reactome pathways for", drug, "\n")
    
  }
}

```

## Combine all enrichments into one master dataframe

```{r}

# Path to your enrichment folder
results_dir <- "/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/results/humanPVATsn/network_analysis/drug_enrichment/step1_key_1000"

# Get list of all CSV files
files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)

# Combine into one dataframe
all_enrichments <- do.call(rbind, lapply(files, function(f) {
  df <- read.csv(f)
  df$Drug_Name <- gsub("_", " ", gsub("^enriched_reactome_|\\.csv$", "", basename(f)))
  return(df)
}))

# Filter by significance
filtered_enrichments <- all_enrichments %>%
  filter(p_value < 1e-12)  # adjust if needed

 # Save as CSV file
write.csv(filtered_enrichments, "/Users/hannahbazin/Desktop/Cambridge/Academics/Han_Lab/MPhil/mphil-project/results/humanPVATsn/network_analysis/drug_enrichment/step1_key_1000/all_enrichments_1e_12.csv", row.names = FALSE)

```





























